Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2023-02-14 analysis/paper/paper.qmd
finished editing discussion section

We further analysed temporal change by looking at specific assemblages in each chronological phase. Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we included CV values of body length, maximum length, mid width, and tang width for assemblages that have five or more stemmed points. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`, with no directional trend apparent between Phase 2 and Phase 3. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage. Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals. It is interesting because SYG1_2 has the strong correlation between attributes (@fig-plot-cor-four-assemblage). Perhaps it could be related to its small sample size, relative to the other assemblages compared here. These results suggest that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.

@@

We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was highly dependent on raw material. We excluded raw materials that were used for less than three points. @fig-plot-cv-by-raw-material shows that quartzite, the raw material for the stemmed points at Sachang, has the highest CV values, and also has wide conference intervals. Other raw materials are generally stable in variation throughout the attributes.

```{r}
#| label: fig-plot-cv-by-raw-material
#| fig-cap: CV values for raw materials that were used for making more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the confidence intervals.
#| fig-width:   8.5
#| fig-height:  4

@@

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). Following @bettinger1999point, we used `r threshold_value_cor` as a threshold of correlation coefficients to evaluate the two transmission biases. Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias. Since there is no clear criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques]. If the CV value is lower than `r threshold_value`, we interpreted this as indicating indirect bias while values higher than `r threshold_value` were interpreted as guided variation.

Our correlation analyses show that the correlation coefficients of stemmed points are all positive but mostly under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias. Some correlations such as body length and maximum length have strong relationship while other correlations such as tang length and body length are much weaker. The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but there is no clear pattern and the values vary depending on assemblages (@fig-plot-cor-four-assemblage). The CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found confirmation of minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points. We found that each assemblage has different distribution patterns of correlation coefficients and CV values. SYG1_2 and SYG1_3 are from the same site but different chronological periods and correlations decreases over time. It is likely that correlation or overall shape can be more depending on individual knappers. We also found that some attributes such as maximum width and maximum length have stronger correlation while the others including stemmed length and body length have lower or even negative correlation. We assume that certain attributes were more carefully transmitted than others. We observed that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes. The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations. We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact. These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.

The most striking differences in both correlation coefficients and CV values is in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material). All distributions are different and there is no clear pattern, indicating that raw materials strongly influence stemmed points. We expected that stemmed points with higher correlation coefficients would have lower CV values, but the actual results do not match our expectations. For example, artifacts made from acidic quartzite have highest correlation but their CV values are also much higher with wider confidence intervals, than all other raw materials. This conflicting results might be related to the relatively unpredictable knapping quality of quartzite. These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals. This makes is harder for the knapper to repeatedly produce the same size and shape end product. This hints at the possibility that physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability. However with only three artifacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.

>>>>

We further analysed temporal change by looking at specific assemblages in each chronological phase. Following previous studies [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we included CV values of body length, maximum length, mid width, and tang width for assemblages that have five or more stemmed points. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`, with no directional trend apparent between Phase 2 and Phase 3. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage. Among the assemblages, SYG1_2 has higher CV values as well as the widest ranges of confidence intervals. It is interesting because SYG1_2 has the strong correlation between attributes (@fig-plot-cor-four-assemblage). Perhaps it could be related to its small sample size, relative to the other assemblages compared here. The larger assemblages in @fig-plot-cv-four-assemblage indicate guided variation, while the smaller assemblages that dominate @fig-plot-cv-over-time suggest indirect bias. Perhaps smaller assemblages represented a social context of higher fidelity copying because the cost of failure was greater due to low social insurance because of low population network sizes. Another interpretation is that these results suggest that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.

@@

We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was dependent on raw material. We excluded raw materials that were used for less than three points. @fig-plot-cv-by-raw-material shows that quartzite, the raw material for the stemmed points at Sachang, has the highest CV values, and also has wide conference intervals. Other raw materials are generally low, right around our threshold value, and stable in variation across the attributes.

```{r}
#| label: fig-plot-cv-by-raw-material
#| fig-cap: CV values for raw materials that were used for making more than three stemmed points. The points represent CV values for each attribute. The vertical lines indicate the confidence intervals.
#| fig-width:   8.5
#| fig-height:  4

@@

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we drew on concepts of cultural transmission. We asked three questions to examine cultural transmission process over time: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and, Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups making stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). Following @bettinger1999point, we used `r threshold_value_cor` as a threshold value for interpreting correlation coefficients to evaluate which of the two transmission biases was dominant. Correlation coefficients below `r threshold_value_cor` are indicative of guided variation and higher values indicate indirect bias. Since there is no previously established criteria for CV, we set roughly `r threshold_value` as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques]. If the CV value is lower than `r threshold_value`, we interpreted this as indicating indirect bias while values higher than `r threshold_value` were interpreted as guided variation.

Our correlation analyses show that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias. Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker. The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage). When all artefacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below. However, further analysis by site and raw material show more ambiguous patterns, with CV values close to or below the threshold value. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points. We found that each assemblage has different distribution patterns of correlation coefficients and CV values. SYG1_2 and SYG1_3 are from the same site but different chronological periods and correlations decreases over time. It is likely that correlation or overall shape can be more depending on individual knappers. We also found that some attributes such as maximum width and maximum length have stronger correlation while the others including stemmed length and body length have lower or even negative correlation. Based on these findings, we propose that certain attributes were more carefully transmitted than others. We observed that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes.

Our interpretation is that attributes with lower CV values were be transmitted in a social context dominated by indirect bias, while the higher CV attributes were influenced more by specific manufacturing and maintenance situations. We assume that the lower CV values of the four attributes imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact. These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.

The most striking differences in both correlation coefficients and CV values is in comparisons of points grouped by raw materials (@fig-corr-raw-ass, @fig-plot-cv-by-raw-material). Raw materials strongly influenced stemmed points shape variance. We expected that stemmed points with higher correlation coefficients would also have lower CV values, but the results do not match our expectations because of how influential raw materials are in the CV values. For example, artifacts made from acidic quartzite have highest correlation but their CV values are also much higher with wider confidence intervals, than all other raw materials. This complex results might be related to the relatively unpredictable knapping quality of quartzite. These grains can redirect knapping forces in unexpected directions, resulting in unintended flake removals. This makes is harder for the knapper to repeatedly produce the same size and shape end product. This hints at the possibility that physical properties of the stone artifact raw materials were a key mechanism in determining shape and size variability. However with only three artifacts made from quartzite in our sample, further analysis is needed to make robust conclusions about the influence of raw materials here.
